---
title: 'EDS 223: assignment 1'
author: 'Elke Windschitl'
date: "2022-09-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## introduction

The following exercises are modified from [Chapter 9 of Geocomputation with R by Rovin Lovelace](https://geocompr.robinlovelace.net/adv-map.html). Each question lists the total number of points. The breakdown of points can be found at the end of each instruction in parentheses. A general grading rubric can be found on the [course website](https://ryoliver.github.io/EDS_223_spatial_analysis/assignments.html).

**Please update "author" to list your first and last names and any collaborators (e.g. Ruth Oliver, Friend1, Friend2)**

**Due by midnight Saturday 2022-10-08**

## prerequisites

```{r load, include=TRUE, message=FALSE, warning=FALSE}
library(sf)
library(raster)
library(dplyr)
library(spData)
library(spDataLarge)
library(tmap)
library(RColorBrewer)
library(grid)
library(leaflet)
```

These exercises rely on a new data object based on the **world** and **worldbank_df** datasets from the \*\*spData\* package.

```{r include=TRUE}
africa = world |> 
  filter(continent == "Africa", !is.na(iso_a2)) |> 
  left_join(worldbank_df, by = "iso_a2") |> 
  dplyr::select(name, subregion, gdpPercap, HDI, pop_growth) |> 
  st_transform("+proj=aea +lat_1=20 +lat_2=-23 +lat_0=0 +lon_0=25")
```

We will also use the **zion** and **nlcd** datasets from the **spDataLarge** package.

```{r include=TRUE}
zion = st_read((system.file("vector/zion.gpkg", package = "spDataLarge")))
data(nlcd, package = "spDataLarge")
force(nlcd)
```

## question 1

##### 10 points

Create a map showing the geographic distribution of the Human Development Index (HDI) across Africa.\
- use both base graphics (hint: use **plot()**) and **tmap**) (4)\
- name two advantages of each based on the experience (3) - name three other mapping packages and an advantage of each (3)

```{r include=TRUE}
# Create a map of the geographic distribution of HDI using plot
plot(africa["HDI"])
     # Not sure what the error is that I'm getting

# Create a map of the geographic distribution of HDI using tmap
tm_shape(africa) +
  tm_polygons(col = "HDI") +
  tm_layout(main.title = "HDI by Countries in Africa",
            main.title.size = 1,
            bg.color = "tan")

```

**Advantages of plot()**

-   The default legend in plot is really well placed and apparent. It makes it easier for the viewer to know what they are looking at.

-   The default colors used in plot have high contrast which make it easy to distinguish which countries fit into which category.

**Advantages of tmap**

-   Tmap feels a lot like ggplot which many r users are highly familiar with. This makes it nice and intuitive when adding layers and elements to the map.

-   While the default colors on tmap are not as contrasted as plot(), they show a nice scale ranging for light to dark of the same colors (red and orange here), and they clearly indicate to the viewer where high or low HDI occur in Africa.

**Three other mapping packages and when you might use them/their advantages:**

-   Use `maps` to outline countries and label cities with points

-   Use `mapview` to create maps that are easily interactive

-   Use `ggplot2` to create basic maps especially with points made up of lat,long data

## question 2

##### 20 points

Extend the **tmap** created for question 1 so the legend has three bins: "high" (HDI above 0.7), "medium" (HDI between 0.55 and 0.7), and "low" (HDI below 0.55). (5)\
- change the legend title (5)\
- change the class labels (5)\
- change the color palette (5)

```{r include=TRUE}
# Add to the map
hdi_africa <- tm_shape(africa) +
  tm_polygons(col = "HDI",
              breaks = c(0, 0.55, 0.7, 0.75),
              labels = c("Low", "Medium", "High"),
              title = "Human Density Index",
              palette = brewer.pal(n = 4, name = "Blues"),
              contrast = 1.7) +
  tm_layout(main.title = "HDI by Countries in Africa",
            main.title.size = 0.8,
            bg.color = "snow2",
            legend.position = c(0.01, 0.01))
hdi_africa
```

## question 3

##### 20 points

Represent Africa's subregions on the map. (5)\
- change the color palette (5)\
- change the legend title (5)\
- combine this map with the map from question 2 into a single plot (5)

```{r include=TRUE}
# Create a map displaying the subregions of Africa
sub_africa <- tm_shape(africa) +
  tm_polygons(col = "subregion",
              palette = brewer.pal(n = 5, name = "Spectral"),
              alpha = 0.9,
              title = "Subregion Name") +
  tm_layout(main.title = "Subregions of Africa",
            main.title.size = 0.8,
            bg.color = "snow2",
            legend.position = c(0.01, 0.01))
sub_africa


# Place the two maps side by side in one plot
tmap_arrange(hdi_africa, sub_africa) 
```

## question 4

##### 30 points

Create a land cover map of Zion National Park (5)\
- change the default colors to match your perception of land cover categories (5)\
- move the map legend outside of the map to improve readability (5)\
- add a scale bar and north arrow and change the position of both to improve the maps aesthetics (5)\
- add the park boundaries on top of the land cover map (5)\
- add an inset of Zion's location in the context of the state of Utah (5)\
- hint: an object representing Utah can be subset from the **us_states** dataset)

```{r include=TRUE}
# First plot the zion polygon
zi_map <- tm_shape(zion) +
  tm_borders(lwd = 2,
             col = "black")

# Create a color palette that makes sense for landcover
land_palette <- c("#33A2FF", "#B2B6B9", "#E9E3C8", 
             "#1D5F2A", "#A9E18E", "#E154CE", 
             "#58360F", "#265364")
# Add the land cover data
lc_map <- tm_shape(nlcd) +
  tm_raster(alpha = 1.0,
            palette = land_palette,
            title = "Land Type") +
  tm_layout(legend.outside = TRUE, # Moving the legend
            main.title = "Land Cover in Zion NP",
            main.title.size = 1) + 
  tm_compass(type = "4star", # Adding the compass
             position = c("0.02", "0.05"),
             size = 4) +
  tm_scale_bar(breaks = c(0, 2, 4, 6, 8, 10), # Adding the scale bar
               position = c("right", "top"),
               bg.color = "white") +
  tm_graticules() +
  zi_map

# Create the boundaries for the inset map
zion$geom
box <- st_bbox(c(xmin = 302903.1, xmax = 334735.5,
               ymin = 4112244, ymax = 4153087),
               crs = st_crs(zion)) %>% 
  st_as_sfc()

# Double check your map of interest
lc_map

# Make the broader map of Utah
ut_geom <- us_states %>% 
  filter(NAME == "Utah")
ut_map <- tm_shape(ut_geom) +
  tm_polygons() +
  tm_shape(zion) + 
  tm_borders(col = "black") +
  tm_shape(box) +
  tm_borders(col = "black")
ut_map 

# Merge the maps
lc_map
print(ut_map, vp = viewport(0.8, 0.27, width = 0.5, height = 0.5))

# I don't feel like the aesthetics look amazing, but I spent a long time to get them to this point!
```

## question 5

##### 20 points

Create an interactive map of the world (10)\
- include a legend (5)\
- change the color palette (5)\
- bonus: use **leaflet** insted of **tmap** (2)

```{r include=TRUE}
# Use tmap to make an interactive map of the world with a legend of the continents
tmap_mode("view")
tm_shape(world) +
  tm_polygons(col = "continent",
              palette = "RdYlBu")

# Try using leaflet to make the map
leaf_map <- leaflet(world) %>% addTiles()
leaf_map

factpal <- colorFactor("RdYlBu", world$continent, n = 8)
continents <- unique(world$continent)

leaflet(world) %>%
  addPolygons(stroke = FALSE, smoothFactor = 0.2, fillOpacity = 1,
              color = ~factpal(continent))

  # addLegend(colors = factpal, 
  #           values = continents,
  #           opacity = 1, 
  #           position = "topright")

# Okay I really can't figure out how to get a legend in here :/

```

## bonus question

##### 5 points

Create THE WORST map! AKA a map that breaks all of the rules of legibility/clarity, yet somehow still passes for a map. We'll vote on the best worst map (worst worst map?) in class.

```{r include=TRUE}
# This is the map (plus a few edits) I originally thought we were supposed to make for question #3, and I thought it looked terrible. Luckily, I figured out #3 nd now here is my terrible-looking map :)
tmap_mode("plot")
# Recreate map #2 with a gray scale
gray_hdi <- tm_shape(africa) +
  tm_polygons(col = "HDI",
              breaks = c(0, 0.55, 0.7, 0.75),
              labels = c("Low", "Medium", "High"),
              title = "Human Density Index",
              palette = brewer.pal(n = 4, name = "Greys"),
              contrast = 1.7) +
  tm_layout(main.title = "HDI by Countries in Africa",
            main.title.size = 1,
            bg.color = "snow2")

# Recreate map #3 with a low alpha to prep for stacking
trans_sub <- sub_africa <- tm_shape(africa) +
  tm_polygons(col = "subregion",
              palette = brewer.pal(n = 5, name = "Spectral"),
              alpha = 0.2) +
  tm_layout(main.title = "Subregions of Africa",
            main.title.size = 1,
            bg.color = "navy") +
  tm_borders(col = "subregion")

gray_hdi + 
  trans_sub +
  tm_layout(main.title = "HDI in subregions of Africa")
```
